Hippocampal replay of experience at real-world speeds

### 2021-07-28 elife_revision1.tex
Add diff

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{YuDistincthippocampalcorticalmemory2017}, although another report suggested that events that represent a single location are only seen in young animals \citep{FarooqEmergencepreconfiguredplastic2019}. Thus, while it would seem useful to replay memories associated with single locations or a small region of the environment, the prevalence of that type of replay in awake animals remains unclear.

@@

We note here that the choice of any particular threshold for classifying the movement dynamic of a SWR is arbitrary, and that the power of our approach lies in part on the ability to assign a probability for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.80 was to use a threshold that corresponds to highly probable events and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold often leads to the conclusion that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verify that our central results hold with a higher threshold of 0.95. We do not know if downstream neurons can explicitly responds to these dynamics based on either threshold (there is probably no hard boundary between these dynamics), but the dynamics correspond to fundamentally different sets of speeds and can be used as a convenient tool in order to understand and summarize the SWR.

@@

\subsection*{Most replays contain coherent spatial content}
After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}; position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal. Even including theta sequences and replay sequences, which can represent positions other than where the animal is, we found that the average median difference between actual and decoded location is small (7 cm median difference, 5-9 cm 95\% CI, for all times where the animal was moving greater than 4 cm/s, 5 fold cross validation). This level of accuracy is just above the size of our 3 cm position bins, and is comparable to that of other studies \citep{DavidsonHippocampalReplayExtended2009, ShinDynamicsAwakeHippocampalPrefrontal2019, FarooqEmergencepreconfiguredplastic2019}. Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during movement, demonstrating the power of our algorithm. For each animal, we also verified that the population spiking rate was relatively consistent over all positions on the track, suggesting that we have comparable sampling of all locations in the environment (\FIGSUPP[2]{fig2sup3}).

@@

We repeated our classification analysis with a higher classification threshold of 0.95 to ensure that our result was not dependent on the threshold of 0.80. We find that, while this change slightly reduced the total fraction of classified SWRs (19,478 of 26,159 or 74\% of all SWRs), an even higher fraction of the classified SWRs (19,317 of 19,478 or 99\% classified SWRs) included spatially coherent content. Similarly, SWRs containing spatially incoherent content comprised a small fraction of the classified SWRs (490 of 19,478 or 3\% classified SWRs). This shows that

@@

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, but instead correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (14,989 of 21,433 or 70\% of classified SWRs) were classified as having only a single dynamic or dynamic mixture (\FIG{5}A, for specific example SWR see \FIG{3}B), whereas SWRs with multiple dynamics or dynamic mixtures (such as those in \FIG{3}A, C, D, E, and F) were less common. Interestingly, the most common category of SWR were those with only stationary-continuous mixtures (8,944 of 21,433 or 42\% of classified SWRs, \FIG{5}A, \FIGSUPP[5]{fig5sup1}). These events contain representations that move at slower speeds (\FIG{5}D)---similar to those speeds experienced by the animal in the environment (\FIG{5}G, median 17 cm/s for run periods and 4 cm/s for all times)---and are sustained, but slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 73 ms vs. continuous 94 ms, \FIG{5}B). Nonetheless, both the slow-moving events and continuous events frequently represent locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 52 cm vs. continuous 43 cm, \FIG{5}C). This indicates that the content of these SWRs, like those that are classified as continuous, do not represent the animal's actual position. We note that roughly a third of these events persisted for the entire duration of the SWR, but the other two thirds included some time where the model was uncertain about the dynamic. This shows the power of our state space decoder because it allows us identify periods of time when the model has high confidence with each SWR, and use these periods to characterize the SWR overall or to focus on specifically for further analysis.

@@

We then carried out a series of control analyses to determine whether our results could be due to differences in spiking statistics across dynamics or interneuron-specific activity patterns. We first calculated the multiunit spiking rates and found that these were similar across the different dynamics. (\FIG{5}F). This indicates that all dynamics, including stationary and stationary-continuous mixtures, were driven by comparable levels of sustained spiking information and could not be explained by the absence of spiking. Further corroborating this, we found that when we re-ran our analysis using periods of high multiunit activity instead of SWRs, we still found a larger proportion of stationary and stationary-continuous dynamics relative to the continuous dynamics \FIGSUPP[5]{fig5sup3}F).

@@

% Figure Data
\figdata{Table of replay statistics for each SWR for Figure 6.}

% Figure 6-Supplemental 1
\figsupp[Examples of fits from the standard "Bayesian" decoder and state space model.]{Examples of fits from the standard "Bayesian" decoder and state space model. (A-F) For each panel, the top row is a raster plot of the sorted cells, the second row is the decoded posterior probability of position in 20 ms time bins for the sorted spikes and the corresponding Radon line fit (blue line) and the MAP fit (green line) on this posterior probability, the third row is a raster plot of the multiunit spike times used for clusterless decoding, the fourth row is the posterior probability in 20 ms time bins using clusterless data as well as the accompanying Radon fit and MAP fit, the fifth row is the posterior probability from the clusterless state space model, and the sixth row is the probability of each dynamic (lines) and the color shading indicates the category of dynamic as before.  }{\includegraphics[width=1.0\linewidth]{figures/Figure6-supplemental1/Figure6-supplemental1_final}}\label{figsupp:fig6sup1}

Finally, we asked how our method compares to the standard approaches to identifying significant replay events via a non-parametric shuffle test and characterizing their content. This comparison is complicated due to the variety of parameters and inference methods used previously, so here we chose three commonly used approaches and a reasonable parameter set for the comparison. We began with an examination of events based on statistical significance of a line fit to a circularly permuted spatial distribution, which has been the standard in the field.  We applied the clusterless algorithm in 20 ms bins to produce probability densities of represented locations, as this binning is comparable to previous methods. We then used the Radon transform method, which finds the line with the highest summed posterior probability of position along the line \citep{DavidsonHippocampalReplayExtended2009, FarooqEmergencepreconfiguredplastic2019}, to calculate, for each SWR, a measure of significance based on the assumption that the probability density moves at a constant speed throughout the event (see examples in \FIGSUPP[6]{fig6sup1}A-F, fourth row, blue line). We found that \texttildelow12\% of the classified events (as defined by our state-space decoder) were identified as significant by this Radon method. Not surprisingly, the fraction of significant events was strongly related to which dynamic they contained. Of the SWRs that contained a period of continuous dynamics, 31\% (1,386 of 4,511 SWRs) were significant. Of all SWRs that contained a period of stationary dynamics, 8\% (542 of 7,176 SWRs) were significant, and when we restricted the analysis to long stationary events (>100 ms duration), we found that 28\% were significant. This increase is not surprising, since the shuffle used for the Radon method penalizes shorter events. An alternative approach, the linear regression method, which samples from the posterior probability and uses linear regression to fit a line to the samples \citep{KarlssonAwakereplayremote2009, CarrHippocampalreplayawake2011, ShinDynamicsAwakeHippocampalPrefrontal2019}, detects only non-zero slopes as significant, and thus by definition cannot identify stationary events as significant.

Conclusions about the representational content of events were similarly variable across methods. In addition to the Radon and linear regression methods, we also computed the MAP estimate, the maximum probability position of each 20 ms time bin (see examples in \FIGSUPP[6]{fig6sup1}A-F, fourth row, green line, \cite{PfeifferAutoassociativedynamicsgeneration2015, StellaHippocampalReactivationRandom2019, KaeferReplayBehavioralSequences2020}). We then compared the speed of the replay for each dynamic and for the entire SWR (\FIG{6}). We found that the methods that rely on fitting a line for the entire SWR (the Radon transform and the linear regression), tend to estimate that the replays proceed at faster speeds (\FIG{6}A,C), whereas the MAP method (\FIG{6}B), which only relies on picking the most probable position for a given time bin and allows for more variable speeds, can capture both fast and slow speeds. In addition, the speeds from the MAP method (\FIG{6}B) are quite similar to the speeds estimated by the state space decoder (\FIG{6}D). This is to be expected because the MAP method imposes less smoothing (although some is added by using 20 ms time bins) whereas the linear methods impose a larger amount of smoothing and cannot change directions (see \FIGSUPP[6]{fig6sup1}A for an example where this is beneficial because the trajectory is linear and the smoothing ignores a lack of spiking and \FIGSUPP[6]{fig6sup1}B for an example where this is not beneficial because the linear fit cannot change direction). Our state space decoder imposes much less smoothing than the linear methods and therefore is closer to the MAP estimates (\FIGSUPP[6]{fig6sup1}B), but also takes into account the uncertainty of each time bin (\FIGSUPP[6]{fig6sup1}C). These results indicate that the methods that made more assumptions about the structure of replay (Radon and linear regression approaches) yield different conclusions about which events are potentially meaningful and how those events progress over time from methods that make fewer assumptions (the MAP and state-space approaches). The substantial variability in both the proportions of "significant" events and the speed of the events demonstrates that key conclusions about the events could be very different depending on the specific method and parameters used.

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (<100 cm/s), which is consistent with the speeds that rats experience their environment. The next most common category were stationary events \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated a persistent representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events  with faster continuous dynamics (\texttildelow50-100 ms), consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics most often represented the animal's current location, similar to continuous events identified in previous work \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}; however, slower events were also capable of representing locations far from the animal. Interestingly, Louie and Wilson \citeyearpar{LouieTemporallyStructuredReplay2001} found that during REM sleep, the longer time scale population activity (on the order of seconds) is highly correlated with firing during times when the animal is running, resembling the animal's experience. Our results are similar but on the timescale of SWRs, which last 10s to 100s of milliseconds, and during awake behavior.

@@

Fourth, our model avoided using large temporal bins and does not make assumptions about the linear progression of events. Instead, the model allows for movement in any direction allowed by geometry of the environment, and further, still accounts for the uncertainty of the data. Because the model gives a moment-by-moment estimate of the position and dynamic, this also minimizes the effect of misidentifying the SWR start and end times. Combined with an acausal estimation of latent position, this approach allows us to identify the specific timing of transitions in dynamics at a fine timescale. In addition to our SWR-based analyses, this approach could also be used to decode activity during movement, where we might expect that place-specific activity would express combinations of stationary and continuous dynamics reflecting different speeds of representational movement during each cycle of the theta rhythm. Similarly, because our approach uses the same encoder-decoder approach as the standard decoder, it can be used to investigate sequence dynamics before experience \cite{DragoiDistinctpreplaymultiple2013}.

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.80 to classify the probability of each state $Pr(I_{k} \mid O_{1:T})$ into 5 categories. Time periods during sharp wave ripples are labeled as stationary, continuous, or fragmented when the probability of each individual state is above 0.80, regardless of duration. Time periods are labeled as stationary-continuous-mixture or fragmented-continuous-mixture when the sum of stationary and continuous or fragmented and continuous are above 0.80, respectively. Time periods where none of these criterion are met are considered unclassified.

>>>>

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{YuDistincthippocampalcorticalmemory2017}, although another report suggested that events that represent a single location are only seen in young animals \citep{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations or a small region of the environment, the prevalence of that type of replay in awake animals remains unclear.

@@

We note here that the choice of any particular threshold for classifying the movement dynamic of a SWR is arbitrary, and that the power of our approach lies in part on the ability to assign a probability for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.80 was to use a threshold that corresponds to highly probable events and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold often leads to the conclusion that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verify that our central results hold with a higher threshold of 0.95.

@@

\subsection*{Most replays contain coherent spatial content}
After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}; position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal. Even including theta sequences and replay sequences, which can represent positions other than where the animal is, we found that the average median difference between actual and decoded location is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation). This level of accuracy is just above the size of our 3 cm position bins, and is comparable to that of other studies \citep{DavidsonHippocampalReplayExtended2009, ShinDynamicsAwakeHippocampalPrefrontal2019, FarooqEmergencepreconfiguredplastic2019}. Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during movement, demonstrating the power of our algorithm. For each animal, we also verified that the population spiking rate was relatively consistent over all positions on the track, suggesting that we have comparable sampling of all locations in the environment (\FIGSUPP[2]{fig2sup3}).

@@

We repeated our classification analysis with a higher classification threshold of 0.95 to ensure that our result was not dependent on the threshold of 0.80. We find that, while this change slightly reduced the total fraction of classified SWRs (19,478 of 26,159 or 74\% of all SWRs), an even higher fraction of the classified SWRs (19,317 of 19,478 or 99\% classified SWRs) included spatially coherent content. Similarly, SWRs containing spatially incoherent content comprised a small fraction of the classified SWRs (490 of 19,478 or 3\% classified SWRs).

@@

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, but instead correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (14,989 of 21,433 or 70\% of classified SWRs) were classified as having only a single dynamic or dynamic mixture (\FIG{5}A, for specific example SWR see \FIG{3}B), whereas SWRs with multiple dynamics or dynamic mixtures (such as those in \FIG{3}A, C, D, E, and F) were less common. Interestingly, the most common category of SWR were those with only stationary-continuous mixtures (8,944 of 21,433 or 42\% of classified SWRs, \FIG{5}A, \FIGSUPP[5]{fig5sup1}). These events contain representations that move at slower speeds (\FIG{5}D)---similar to those speeds experienced by the animal in the environment (\FIG{5}G)---and are sustained, but slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 73 ms vs. continuous 94 ms, \FIG{5}B). Nonetheless, both the slow-moving events and continuous events frequently represent locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 52 cm vs. continuous 43 cm, \FIG{5}C). This indicates that the content of these SWRs, like those that are classified as continuous, do not represent the animal's actual position. We note that roughly a third of these events persisted for the entire duration of the SWR, but the other two thirds included some time where the model was uncertain about the dynamic. This shows the power of our state space decoder because it allows us identify periods of time when the model has high confidence with each SWR, and use these periods to characterize the SWR overall or to focus on specifically for further analysis.

@@

We then carried out a series of control analyses to determine whether our results could be due to differences in spiking statistics across dynamics or interneuron-specific activity patterns. We first calculated the multiunit spiking rates and found that these were similar across the different dynamics. (\FIG{5}F). This indicates that all dynamics, including stationary and stationary-continuous mixtures, were driven by comparable levels of sustained spiking information and could not be explained by the absence of spiking.

@@

\figdata{Table of replay statistics for each SWR for Figure 6.}
\end{figure}

Finally, we asked how our method compares to the standard approaches to identifying significant replay events via a non-parametric shuffle test and characterizing their content. This comparison is complicated due to the variety of parameters and inference methods used previously, so here we chose three commonly used approaches and a reasonable parameter set for the comparison. We began with an examination of events based on statistical significance of a line fit to a circularly permuted spatial distribution, which has been the standard in the field.  We applied the clusterless algorithm in 20 ms bins to produce probability densities of represented locations, as this binning is comparable to previous methods. We then used the Radon transform method, which finds the line with the highest summed posterior probability of position along the line \citep{DavidsonHippocampalReplayExtended2009, FarooqEmergencepreconfiguredplastic2019}, to calculate, for each SWR, a measure of significance based on the assumption that the probability density moves at a constant speed throughout the event. We found that \texttildelow12\% of the classified events (as defined by our state-space decoder) were identified as significant. Not surprisingly, the fraction of significant events was strongly related to which dynamic they contained. Of the SWRs that contained a period of continuous dynamics, 31\% (1,386 of 4,511 SWRs) were significant. Of all SWRs that contained a period of stationary dynamics, 8\% (542 of 7,176 SWRs) were significant, and when we restricted the analysis to long stationary events (>100 ms duration), we found that 28\% were significant. This increase is not surprising, since the shuffle used for the Radon method penalizes shorter events. An alternative approach, the linear regression method, which samples from the posterior probability and uses linear regression to fit a line to the samples \citep{KarlssonAwakereplayremote2009, CarrHippocampalreplayawake2011, ShinDynamicsAwakeHippocampalPrefrontal2019}, detects only non-zero slopes as significant, and thus by definition cannot identify stationary events as significant.

Conclusions about the representational content of events were similarly variable across methods. In addition to the Radon and linear regression methods, we also computed  the MAP estimate, the maximum probability position of each 20 ms time bin \citep{PfeifferAutoassociativedynamicsgeneration2015, StellaHippocampalReactivationRandom2019, KaeferReplayBehavioralSequences2020}. We then compared the speed of the replay for each dynamic and for the entire SWR (\FIG{6}). We found that the methods that rely on fitting a line for the entire SWR (the Radon transform and the linear regression), tend to estimate that the replays proceed at faster speeds (\FIG{6}A,C), whereas the MAP method (\FIG{6}B), which only relies on picking the most probable position for a given time bin and allows for more variable speeds, can capture both fast and slow speeds. In addition, the speeds from the MAP method (\FIG{6}B) are quite similar to the speeds estimated by our decoder (\FIG{6}D). This is to be expected because the MAP method imposes less smoothing (only the 20 ms time bins) whereas the linear methods impose a larger amount of smoothing. Our decoder imposes more smoothing than the MAP estimate but much less than the linear methods and therefore is closer to the MAP estimates. These results indicate that the methods that made more assumptions about the structure of replay (Radon and linear regression approaches) yield different conclusions about which events are potentially meaningful and how those events progress over time from methods that make fewer assumptions (the MAP and state-space approaches). The substantial variability in both the proportions of "significant" events and the speed of the events demonstrates that key conclusions about the events could be very different depending on the specific method and parameters used.

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (<100 cm/s), which is consistent with the speed that rats move through their environment. The next most common category were stationary events \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated a persistent representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events  with faster continuous dynamics (\texttildelow50-100 ms), consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics most often represented the animal's current location, similar to continuous events identified in previous work \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}; however, slower events were also capable of representing locations far from the animal.

@@

Fourth, our model avoided using large temporal bins and does not make assumptions about the linear progression of events. Instead, the model allows for movement in any direction allowed by geometry of the environment, and further, still accounts for the uncertainty of the data. Because the model gives a moment-by-moment estimate of the position and dynamic, this also minimizes the effect of misidentifying the SWR start and end times. Combined with an acausal estimation of latent position, this approach allows us to identify the specific timing of transitions in dynamics at a fine timescale. In addition to our SWR-based analyses, this approach could also be used to decode activity during movement, where we might expect that place-specific activity would express combinations of stationary and continuous dynamics reflecting different speeds of representational movement during each cycle of the theta rhythm.

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.80 to classify the probability of each state $Pr(I_{k} \mid O_{1:T})$ into 5 categories. Time periods during sharp wave ripples are labeled as stationary, continuous, or fragmented when the probability of each individual state is above 0.80. Time periods are labeled as stationary-continuous-mixture or fragmented-continuous-mixture when the sum of stationary and continuous or fragmented and continuous are above 0.80, respectively. Time periods where none of these criterion are met are considered unclassified.
